Methods for Large Scale Learning to Rank

### 2014-09-17 report/Chapters/Implementation.tex
ranking accuracy measurements for validation

\section{ListNet}
The following sections describe the Pig jobs that form the three independent parts of the ListNet ranking model: 1) Preprocessing, 2) Training (this includes evaluation over the validation set) and 3) Testing (evaluation over the test set). 
\subsection{Preprocessing}
The preprocessing phase consists of two separate Pig jobs. The first Pig job determines the minimum and the maximum values per feature in the training set. The second Pig job rescales each feature of the train, validation and test datasets using the following formula for rescaling:
\begin{equation}
x^{'} = \frac{x-min(x)}{max(x)-min(x)}
\end{equation}
This rescaling procedure sets the values of all features to be within range $[0,1]$.\\

The first Pig job:\\
\begin{minipage}{\linewidth}
\begin{lstlisting}
REGISTER [path prefix]/lib/*.jar;
TRAIN = LOAD '[path prefix]/input/[dataset name]/Fold[fold number]/train.txt' USING PigStorage(' ');
TRAIN_STD = FOREACH TRAIN GENERATE flatten(udf.util.ToStandardForm(*));
TRAIN_STD_BY_QUERY = GROUP TRAIN_STD BY $1 PARALLEL [available reducers];
MIN_MAX = FOREACH TRAIN_STD_BY_QUERY GENERATE flatten(udf.util.GetMinMax(*));
MIN_MAX_GRPD = GROUP MIN_MAX ALL;
MIN_MAX_FIN = FOREACH MIN_MAX_GRPD GENERATE flatten(udf.util.CombineMinMax(*));
STORE MIN_MAX_FIN INTO 'minmax[fold number]';
\end{lstlisting}
\end{minipage}\\

The minimum and maximum values per feature stored in MIN\_MAX\_FIN can now be read from \ac{HDFS} and used within the Java framework by using the Windows Azure Storage library. The second Pig job uses these minimum and maximum values by parametrising the \ac{UDF} that performs the rescaling transformation, udf.util.ScaleFeatures(), with the minimum and maximum feature values.\\

\begin{table}
\centering
\begin{tabular}{p{6cm}p{7cm}}\toprule
UDF & Description \\
\midrule
udf.util.ToStandardForm() & Transforms the data set into the standard form of relevance label in first column followed by feature values. Strips data of any other columns, if present.\\
udf.util.GetMinMax() & Extracts the minimum and maximum value per feature, for the documents of a single query.\\
udf.util.CombineMinMax() & Combines outputs of the udf.util.GetMinMax() UDF for each query into globally minimum and maximum feature values.\\
\bottomrule
\end{tabular}
\caption{Description of preprocessing User Defined Functions (1/2)}
\label{tbl:preprocessing_udfs_1}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{p{5cm}p{8cm}}\toprule
UDF & Description \\
\midrule
udf.util.ToStandardForm() & See Table \ref{tbl:preprocessing_udfs_1} for description.\\
udf.util.ScaleFeatures() & Uses the minimum and maximum feature values with which it is parametrised perform the following rescaling transformation to the features: $x^{'} = \frac{x-min(x)}{max(x)-min(x)}$.\\
\bottomrule
\end{tabular}
\caption{Description of preprocessing User Defined Functions (2/2)}
\label{tbl:preprocessing_udfs_2}
\end{table}

\subsection{Training}
The training stage, like the preprocessing stage, consists of two separate Pig jobs. The first Pig job calculates the Cross Entropy loss on training data of the current model and calculates the gradients for the next model update. The second Pig job is an internal validation step that validates the model on the validation set by calculating \ac{nDCG}@k.\\

@@

\begin{table}
\centering
\begin{tabular}{p{6cm}p{7cm}}\toprule
UDF & Description \\
\midrule
udf.listnet.QueryLossGradient() & Calculates the Cross Entropy loss for a query and calculates the gradients per feature based on this query.\\
udf.listnet.ExpRelOurScores() & Calculates the predicted relevance label of a query based on the current model weights and transforms this following the transformation $x -> e^{x}$. In case the current iteration is the first iteration, the same transformation is applied to the ground truth relevance label.\\
udf.listnet.MultiSum() & Calculates aggregated loss and feature gradients by summing the per-query losses and per-query feature gradients.\\
\bottomrule
\end{tabular}
\caption{Description of training User Defined Functions (1/2)}
\label{tbl:training_udfs_1}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{p{6cm}p{7cm}}\toprule
UDF & Description \\
\midrule
udf.util.Ndcg() & Calculates \ac{nDCG}@k for a query.\\
\bottomrule
\end{tabular}
\caption{Description of training User Defined Functions (2/2)}
\label{tbl:training_udfs_2}
\end{table}

>>>>

\section{ListNet}
The following sections describe the Pig jobs that jointly form the three independent parts of the ListNet ranking model: 1) Preprocessing, 2) Training (this includes evaluation over the validation set) and 3) Testing (evaluation over the test set). 
\subsection{Preprocessing}
The preprocessing phase consists of two separate Pig jobs. The first Pig job determines the minimum and the maximum values per feature in the training set. The second Pig job rescales each feature of the train, validation and test datasets using the following formula for rescaling:
\begin{equation}
x^{'} = \frac{x-min(x)}{max(x)-min(x)}
\label{eq:rescaling}
\end{equation}
This rescaling procedure sets the values of all features to be within range $[0,1]$.\\

The first Pig job:\\
\begin{minipage}{\linewidth}
\begin{lstlisting}
REGISTER [path prefix]/lib/*.jar;
TRAIN = LOAD '[path prefix]/input/[dataset name]/Fold[fold number]/train.txt' USING PigStorage(' ');
TRAIN_STD = FOREACH TRAIN GENERATE flatten(udf.util.ToStandardForm(*));
TRAIN_STD_BY_QUERY = GROUP TRAIN_STD BY $1 PARALLEL [available reducers];
MIN_MAX = FOREACH TRAIN_STD_BY_QUERY GENERATE flatten(udf.util.GetMinMax(*));
MIN_MAX_GRPD = GROUP MIN_MAX ALL;
MIN_MAX_FIN = FOREACH MIN_MAX_GRPD GENERATE flatten(udf.util.CombineMinMax(*));
STORE MIN_MAX_FIN INTO 'minmax[fold number]';
\end{lstlisting}
\end{minipage}\\
All expressions between square brackets in the code snippet above are dynamically set by the framework. The minimum and maximum values per feature stored in MIN\_MAX\_FIN are read from storage by the framework after completion of the MapReduce job. The framework passes these minimum and maximum values to the second Pig job by passing an array of minimum and maximum values to the constructor of the udf.util.ScaleFeatures() \ac{UDF}, that performs the rescaling operation of Equation \ref{eq:rescaling}s.\\

\begin{table}
\centering
\begin{tabular}{p{6cm}p{7cm}}\toprule
UDF & Description \\
\midrule
udf.util.ToStandardForm() & Transforms the data set into the standard form of relevance label in first column followed by feature values. Strips data of any other columns, if present.\\
udf.util.GetMinMax() & Extracts the minimum and maximum value per feature, for the documents of a single query.\\
udf.util.CombineMinMax() & Combines outputs of the udf.util.GetMinMax() UDF for each query into globally minimum and maximum feature values.\\
\bottomrule
\end{tabular}
\caption{Description of preprocessing phase User Defined Functions (Pig job 1)}
\label{tbl:preprocessing_udfs_1}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{p{5cm}p{8cm}}\toprule
UDF & Description \\
\midrule
udf.util.ToStandardForm() & See Table \ref{tbl:preprocessing_udfs_1} for description.\\
udf.util.ScaleFeatures() & Uses the minimum and maximum feature values with which it is parametrised perform the following rescaling transformation to the features: $x^{'} = \frac{x-min(x)}{max(x)-min(x)}$.\\
\bottomrule
\end{tabular}
\caption{Description of preprocessing phase User Defined Functions (Pig job 2)}
\label{tbl:preprocessing_udfs_2}
\end{table}

\subsection{Training}
The Pig jobs that form the ListNet training phase form an implementation of the theoretical description of ListNet as described in section \ref{sec:ltrmethods_listnet}. The training stage, like the preprocessing stage, consists of two separate Pig jobs. The first Pig job calculates the Cross Entropy loss on training data of the current model and calculates the gradients for the next model update. The second Pig job is an internal validation step that validates the model on the validation set by calculating \ac{nDCG}@k.\\

@@

The actual neural network weights, passed by the framework as constructor parameter in the script above, are administered in Java code. The per-feature gradients calculated by the Pig job above are read from storage after completion and used to update the neural network by, for each feature, adding the feature gradient multiplied with a step size parameter to the previous feature weight.

\begin{table}
\centering
\begin{tabular}{p{6cm}p{7cm}}\toprule
UDF & Description \\
\midrule
udf.listnet.QueryLossGradient() & Calculates the Cross Entropy loss for a query and calculates the gradients per feature based on this query.\\
udf.listnet.ExpRelOurScores() & Calculates the predicted relevance label of a query based on the current model weights and transforms this following the transformation $x \rightarrow e^{x}$. In case the current iteration is the first iteration, the same transformation is applied to the ground truth relevance label.\\
udf.listnet.MultiSum() & Calculates aggregated loss and feature gradients by summing the per-query losses and per-query feature gradients.\\
\bottomrule
\end{tabular}
\caption{Description of training phase User Defined Functions (Pig job 1)}
\label{tbl:training_udfs_1}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{p{6cm}p{7cm}}\toprule
UDF & Description \\
\midrule
udf.util.Ndcg() & Calculates \ac{nDCG}@k for a query.\\
\bottomrule
\end{tabular}
\caption{Description of training phase User Defined Functions (Pig job 2)}
\label{tbl:training_udfs_2}
\end{table}

### 2014-09-17 report/Chapters/LTRMethods.tex
ranking accuracy measurements for validation

\section{ListNet}
ListNet \cite{Cao2007} is a listwise ranking function whose loss function is not directly related to information retrieval evaluation metrics. ListNet's loss function is defined using a probability distribution on permutations. Probability distributions on permutations have been a research topic within the field of probability theory which has been extensively researched. ListNet uses the Plackett-Luce model \cite{Plackett1975,Luce1959}, which is one of the most well-known permutation probability distributions. The Plackett-Luce model defines the probability of all possible permutations $\pi$, given all document ranking scores $S$, as shown in Equation \ref{eq:plackett_luce}.
\begin{equation}
P(\pi|S) = \prod\limits_{j=1}^{m}\frac{\phi(s_{\pi^{-1}(j)})}{\sum\nolimits_{u=1}^{m}\phi(s_{\pi^{-1}(u)})}
\label{eq:plackett_luce}
\end{equation}
ListNet uses Gradient Descent to optimise a neural network such that its Cross Entropy loss compared to the Plackett-Luce distribution over the training data relevance labels is minimal. Note that some sources, including Liu \cite{Liu2007}, describe ListNet as using \ac{KL divergence} as loss function. \ac{KL divergence} and Cross Entropy are however identical up to an additive constant when comparing distribution $q$ against a fixed reference distribution $p$. This follows from the definition of Cross Entropy Loss, which is shown in Equation \ref{eq:cross_entropy}.
\begin{equation}
H(p,q) = H(p) + D_{KL}(p||q)
\label{eq:cross_entropy}
\end{equation}
where $H(p)$ is the entropy of $p$ and $D_{KL}(p||q)$ is the \ac{KL divergence} of $q$ from $p$.\\

>>>>

\section{ListNet}
\label{sec:ltrmethods_listnet}
ListNet \cite{Cao2007} is a listwise ranking function whose loss function is not directly related to information retrieval evaluation metrics. ListNet's loss function is defined using a probability distribution on permutations. Probability distributions on permutations have been a research topic within the field of probability theory which has been extensively researched. ListNet uses the Plackett-Luce model \cite{Plackett1975,Luce1959}, which is one of the most well-known permutation probability distributions. The Plackett-Luce model defines the probability of all possible permutations $\pi$, given all document ranking scores $S$, as shown in Equation \ref{eq:plackett_luce}.
\begin{equation}
P(\pi|S) = \prod\limits_{j=1}^{m}\frac{\phi(s_{\pi^{-1}(j)})}{\sum\nolimits_{u=1}^{m}\phi(s_{\pi^{-1}(u)})}
\label{eq:plackett_luce}
\end{equation}
ListNet uses Gradient Descent to optimise a neural network such that its Cross Entropy loss compared to the Plackett-Luce distribution over the training data relevance labels is minimal. Note that some sources, including Liu \cite{Liu2007}, describe ListNet as using \ac{KL divergence} as loss function. \ac{KL divergence} and Cross Entropy are however identical up to an additive constant when comparing distribution $q$ against a fixed reference distribution $p$. This follows from the definition of Cross Entropy Loss, which is shown in Equation \ref{eq:cross_entropy}.
\begin{equation}
H(p,q) = H(p) + D_{KL}(p||q)
\label{eq:cross_entropy}
\end{equation}
where $H(p)$ is the entropy of $p$ and $D_{KL}(p||q)$ is the \ac{KL divergence} of $q$ from $p$.\\

### 2014-09-17 report/Chapters/Results.tex
ranking accuracy measurements for validation

This chapter describes and discusses the results of the running time measurements of the cluster and baseline implementations as described in \ref{chap:implementation}.
\section{ListNet}
Figure \ref{fig:listnet_train_time} and Figure \ref{fig:listnet_train_time_log} (which presents the same data with a logarithmic data size axis) show the processing times of a single training iteration using the ListNet implementation included in RankLib library and the cluster implementation described in chapter \ref{chap:implementation} with different numbers of data nodes. The horizontal positions of the measurements are identical between execution method, as they are determined by the collection of datasets used. As previously described in the methodology section, the LETOR 3.0, LETOR 4.0 and MSLR-WEB10/30K datasets are used, supplemented with generated datasets that are multiplications of MSLR-WEB30K. Table \ref{tbl:recap_datasets} describes the datasets used in the experiments.

>>>>

This chapter describes and discusses the results of the running time measurements of the cluster and baseline implementations as described in \ref{chap:implementation}. In addition, ranking accuracy results are included to validate the correctness of the Learning to Rank algorithm implementations.
\section{ListNet}
\subsection{Ranking accuracy}
\begin{table}[!h]
\centering
\begin{tabular}{p{6cm}p{1.6cm}p{1.4cm}r}\toprule
Dataset & RankLib & Cluster & Official evaluation \\
\midrule
OHSUMED (LETOR 3.0 version)     & 0.29          &   TODO & 0.3793 \\
MQ2008      					& 0.3865        &   TODO & 0.4689 \\
MQ2007      					& 0.2873        &   TODO & 0.4767 \\
MSLR-WEB10K 					& 0.1744     	&   TODO & - \\
MSLR-WEB30K 					& 0.1837     	&   TODO & - \\
\bottomrule
\end{tabular}
\caption{\acs{NDCG}@10 performance on the testing set on the first fold}
\label{tbl:accuracy_comparison}
\end{table}

\subsection{Speedup}
Figure \ref{fig:listnet_train_time} and Figure \ref{fig:listnet_train_time_log} (which presents the same data with a logarithmic data size axis) show the processing times of a single training iteration using the ListNet implementation included in RankLib library and the cluster implementation described in chapter \ref{chap:implementation} with different numbers of data nodes. The horizontal positions of the measurements are identical between execution method, as they are determined by the collection of datasets used. As previously described in the methodology section, the LETOR 3.0, LETOR 4.0 and MSLR-WEB10/30K datasets are used, supplemented with generated datasets that are multiplications of MSLR-WEB30K. Table \ref{tbl:recap_datasets} describes the datasets used in the experiments.\\
